NeuroElectro: A Window to the World's Neuron Electrophysiology Data

### 2014-03-18 neuroelectro_frontiers/main.tex
Minor changes befoire resubmission

\begin{table}[!t]
\textbf{\refstepcounter{table}\label{tab:journal_listing} Table \arabic{table}.}
{Statistics of journals represented in the NeuroElectro database. Listing of journals and counts of articles downloaded (articles obtained), articles with published data tables containing neurophysiological information which has been manually validated by an expert curator (validated), and articles which likely contain information in a data table which has not yet been manually curated (not validated). 
Not validated articles are those which have at least 4 algorithmically assigned electrophysiological concepts within a data table.}
\processtable{}
{\begin{tabular}{rrrr}
    \toprule
    Journal & Articles obtained & Validated & Not validated \\
    \midrule
    J. Neurosci. & 19002 & 104   & 560 \\
    J. Neurophysiol. & 12078 & 94    & 555 \\
    J. Physiol. (Lond.) & 10543 & 44    & 235 \\
    Neuroscience & 3035  & 14    & 205 \\
    Eur. J. Neurosci. & 2495  & 7     & 117 \\
    Brain Res. & 3017  & 7     & 146 \\
    Neuron & 1657  & 4     & 43 \\
    Epilepsia & 463   & 2     & 23 \\
    Neurosci. Lett. & 1468  & 2     & 34 \\
    Hippocampus & 208   & 2     & 10 \\
    \botrule
    \end{tabular}}
   {}
\end{table}

We obtained tens of thousands of potentially relevant full article texts directly from publisher websites.  
We first identified potential articles that were likely to contain information relevant to neuron biophysics using the native search functions provided within the journal websites and only downloaded articles containing a specific list of terms including ``input resistance'' and ``resting membrane potential'' (Fig. \ref{text_mining_workflow}).  
This pre-selection step allowed us to identify and download only articles that contained data relevant to our project.  
Upon identifying candidate articles, we then downloaded the full text of each potentially-relevant article as HTML; 
articles downloaded from the publisher Elsevier (e.g., \emph{Neuron and Brain Research}) were downloaded as XML using the provided text-mining API and subsequently converted to HTML.  
We chose to work with HTML (as opposed to PDF or XML) because HTML provides a machine-readable markup of the article's content, allowing us to easily identify relevant elements within the article -- such as data tables and the methods section -- using publicly available HTML-parsing tools (here we used the Beautiful Soup HTML-processing library implemented in Python: \url{http://www.crummy.com/software/BeautifulSoup/bs4/doc/}).  
Furthermore, because HTML is a single semi-structured standard used across publishers, we could write relatively generic HTML-processing algorithms applicable to content published across journals.  
Our focus on using HTML limits us to relatively newer articles - typically those published after 1996 - because before this time most publications are only available as scanned PDF files.  
However, because the rate of publication across the field has grown exponentially, this HTML-available subset constitutes the majority of published neuroscience articles.

@@

When designing our processing algorithms, we parsed data strings from right to left: 
first searching for data entities contained within parentheses, then for entities contained to the right of the $\pm$ term, and finally the remaining term which we assumed to refer to the mean term.  
We found that occasionally data were reported as ``XX (LL - HH)'' -- where LL and HH indicate the lower and upper limits of a data range -- and accounted for these cases similarly.  
We used regular expressions to identify entities such as digits, decimal signs, parentheses, and $\pm$ signs.  
We then converted the individual data elements which were encoded as textual strings of digits to double precision decimal entities before storing these into our database.  
Our focus here was primarily on parsing the data record mean value (i.e., summarizing the properties of a number of recorded neurons), but also extracted and stored the error term and sample size where possible.  
Using these methods, we were able to extract 2176 electrophysiological values for 93 distinct neuron types within 279 articles.

@@

\subsection{Object models and relational database}
We stored extracted data and metadata using a relational database implemented in MySQL (\url{http://dev.mysql.com/doc/refman/5.6/en/}) built from a Python Django object model (\url{https://www.djangoproject.com/}).  
The object model contains classes for a number of fields, such as full article texts, electrophysiological properties, neuron types, synonyms, electrophysiological data values, and experimental metadata (Fig. \ref{db_schema}).  
A useful feature of the relational nature of the database is that it enables linking between classes (e.g., linking between neuron types and electrophysiological properties reported by a single investigator across multiple articles). 
This linking feature facilitates efficient and arbitrary querying of data; 
for example, querying for known electrophysiological data on olfactory bulb mitral cells recorded \emph{in vitro} and published between the dates 2000 and 2004.  
For example, such a feature could be used to assess whether the properties of olfactory bulb mitral cells have changed as a function of time or are dependent upon whether the data are collected \emph{in vitro} or \emph{in vivo}.

@@

\begin{itemize}
\item \url{http://neuroelectro.org/api/1/n/} : Returns a list of all neurons with electrophysiological data indexed in NeuroElectro.
\item \url{http://neuroelectro.org/api/1/nedm/?nlex=sao830368389} : Returns a list of all indexed data on CA1 pyramidal cells (queried using the NeuroLex identifier for CA1 pyramidal cells, \emph{sao830368389}).  
\item \url{http://neuroelectro.org/api/1/nes/?e\_\_name=Input\%20Resistance\&n\_\_name=Cerebellum\%20Purkinje\%20Cell} : Returns a data record composed of the mean, standard deviation, and n, summarizing input resistance measurements from cerebellar Purkinje cells based on all indexed articles in NeuroElectro database.  
Here the database query is performed using the textual strings for the electrophysiological and neuron type concepts.
\end{itemize}

@@

Given our laboratory's focus on olfactory circuits, we conducted a natural experiment to compare the efficacy of biophysical property extraction using these semi-automated methods versus traditional methods which do not make use of algorithmic text-mining as a pre-processing step.  
In a seven-hour curation session, a senior graduate student in our laboratory identified 91 electrophysiological measurements (focusing on resting membrane potential, input resistance, membrane time constant, spike amplitude, spike width, and spike threshold) from 35 articles for 7 olfactory bulb neuron types using only prior knowledge of which articles and investigators were likely to have reported such electrophysiological data.  
In a comparable seven-hour curation session using our semi-automated methods, a single curator (with similar expertise to the first curator) identified 551 electrophysiological measurements from 70 articles across 40 neuron types throughout the nervous system.  
Moreover, this comparison would likely tilt even more in favor of the semi-automated methods had the curators been less familiar with the primary literature.

\subsection{Scalability of current approach}
We note that multiple steps in our approach require manual intervention by an expert curator in order for electrophysiological measurements to be extracted with an acceptably low error rate.  
Namely, an expert curator needs to specify which neuron types are recorded from in each article and where data from the normotypic or ``control'' states of these neurons are textually referenced within a data table.  
Moreover, given the current accuracy of the unsurpervised algorithmic assignment of electrophysiological concepts and experimental metadata (78\% and 50\%, respectively), these also need to be manually validated and corrected and normalized as required by an expert.  
Given the necessity of these manual steps, the scalability of our current approach is limited by our ability to manually curate this information or by our ability to improve the error rate of the automated methods.  
Despite this limitation, our current pipeline is still much faster than a purely manual one.  The methodology could be further improved by addressing falsely matching entities (such as EPSP amplitude in section \ref{sec:ephys_extraction}) by either auto-suggesting alternative matches returned by string-similarity algorithms or by simply adding these concepts to the electrophysiolgical ontology.  
Moreover, these improvements would facilitate formally computing the sensitivity and specificity of these entity recognition methods.

\subsection{Preliminary use of NeuroElectro in scientific work}
The NeuroElectro project is intended to facilitate scientific investigation by providing easy access to large quantities of data about neurons.  
Because the data is machine-readable, we have already begun to conduct several analyses that would not be possible without this resource.  
First, we have begun an investigation of the relationships between neurons defined by the similarity of their electrophysiological properties.  
This information can be used to make predictions about as yet unmeasured properties.  
Second, we have begun to explore the relationship between patterns of gene expression (using both the Allen Brain Atlas \citep{lein_genome-wide_2007} and single cell qPCR approaches) and electrophysiological properties of neurons.  
Third, we have begun automated testing of quantitative neuron models in concert with SciUnit \citep{omar_sciunit_2014}, under the reasonable assumption that these models should be constrained by the available experimental data.  
These projects are described in manuscripts currently in preparation.

\subsection{Extensions and improvements to the current semi-automated algorithms}
Currently, neuron type identification is a critical bottleneck in our approach.  
One potential improvement would be to replace the nonspecific bag-of-words approach we are currently using in favor of a bioNLP classifier-based approach \citep{mccallum_mallet:_2002}.  
Specifically, we propose adapting the named entity recognition methodology used by the WhiteText project for tagging brain regions mentioned in literature \citep{french_automated_2009,french_using_2012} and first identifying spans of text likely to pertain to a neuron type before mapping these textual spans to a individual neuron type within the neuron ontology.  
Moreover, such an approach could easily incorporate common neuron type acronyms and abbreviations \citep{okazaki_building_2006}.

@@

Given the relatively low accuracy of the automated approach to identifying neuron types, there may be several avenues through which this process can be improved. 
For example, we note the automated approach was particularly ineffective when the neuron type investigated within an article was not already described in NeuroLex or when the neuron had an insufficient list of synonyms associated with it.  
The current implementation of NeuroElectro also does not consider common neuron type acronyms (e.g., that olfactory bulb mitral cells are commonly referred to as ``MCs'').  
Adding acronym identification to future iterations will thus likely improve the automated approach \citep{okazaki_building_2006,french_using_2012}.  
Moreover, our current implementation of the bag-of-words algorithm would likely be improved via minor improvements, such as only identifying neurons using the text of the abstract or results and discarding text from the introduction or discussion. 
As neuron identification forms the major bottleneck in the scalability of NeuroElectro due to the requirement for manual curation, we plan to address this bottleneck in future revisions.

@@

Future developments in machine extraction of data from the scientific literature will be of great benefit.  
These should include better semantic understanding of context, ranging from relatively unambiguous notations such as units, to syntax-parsing that relates objects of study to their reported properties in free-form prose.  
Much progress has been made by computer scientists in some of these areas, and more future engagement with their research should enable vastly more data to be extracted from the literature.

However, more widespread inclusion of fundamental electrophysiological measurements in data tables in future publications will strongly facilitate data-sharing and meta-analysis initiatives such as NeuroElectro.  
We believe that, if successful, the use of NeuroElectro will influence the practices of scientists writing papers and reporting results.  
Specifically, we recommend the usage of common standards and definitions for basic physiological measurements \citep{toledo-rodriguez_correlation_2004} and neuron types \citep{ascoli_petilla_2008,larson_neurolex.org:_2013}.  
Moreover, we advocate that, where possible, scientists report more basic physiological data overall and report such data using machine-parsable data tables.  
These recommendations could be made informally by journals (in particular, requested by reviewers during manuscript review) as well as by funding agencies.
This change would make it easier for scientists to find and make use of data collected by others.  
Such a culture shift has the potential to make science function more effectively and efficiently to facilitate discovery.

@@

\bibliographystyle{template/frontiersinSCNS&ENG} % for Science and Engineering articles
\bibliography{references}
\
\section*{Supplemental Data}  
\begin{figure*}[htbp]
\centering
\includegraphics[scale = .4, angle =90]{figures/django_models_v2.jpg}
\caption{Illustration of NeuroElectro relational database schema}
\label{db_schema}
\end{figure*}
%SDB: anyway you can increase the font on at least some of these pathways? Or split into multiple figures showing subsections of the relations? As is, it is very difficult to look at this and take anything away from it.

>>>>

\begin{table}[!t]
\textbf{\refstepcounter{table}\label{tab:journal_listing} Table \arabic{table}.}
{Statistics of journals represented in the NeuroElectro database. Listing of journals and counts of articles downloaded (articles obtained), articles with published data tables containing neurophysiological information which has been manually validated by an expert curator (validated), and articles which likely contain information in a data table which has not yet been manually curated (not validated). 
Not validated articles are those which have at least 4 algorithmically assigned electrophysiological concepts within data tables.}
\processtable{}
{\begin{tabular}{rrrr}
    \toprule
    Journal & Articles obtained & Validated & Not validated \\
    \midrule
    J. Neurosci. & 19002 & 104   & 560 \\
    J. Neurophysiol. & 12078 & 94    & 555 \\
    J. Physiol. (Lond.) & 10543 & 44    & 235 \\
    Neuroscience & 3035  & 14    & 205 \\
    Eur. J. Neurosci. & 2495  & 7     & 117 \\
    Brain Res. & 3017  & 7     & 146 \\
    Neuron & 1657  & 4     & 43 \\
    Epilepsia & 463   & 2     & 23 \\
    Neurosci. Lett. & 1468  & 2     & 34 \\
    Hippocampus & 208   & 2     & 10 \\
    \botrule
    \end{tabular}}
   {}
\end{table}

We obtained tens of thousands of potentially relevant full article texts directly from publisher websites.  
We first identified potential articles that were likely to contain information relevant to neuron biophysics using the native search functions provided within the journal websites and only downloaded articles containing in their full text any of a specific list of terms including ``input resistance'' and ``resting membrane potential'' (Fig. \ref{text_mining_workflow}).  
This pre-selection step allowed us to identify and download only articles that contained data relevant to our project.  
Upon identifying candidate articles, we then downloaded the full text of each potentially-relevant article as HTML; 
articles downloaded from the publisher Elsevier (e.g., \emph{Neuron and Brain Research}) were downloaded as XML using the provided text-mining API and subsequently converted to HTML.  
We chose to work with HTML (as opposed to PDF or XML) because HTML provides a machine-readable markup of the article's content, allowing us to easily identify relevant elements within the article -- such as data tables and the methods section -- using publicly available HTML-parsing tools (here we used the Beautiful Soup HTML-processing library implemented in Python: \url{http://www.crummy.com/software/BeautifulSoup/bs4/doc/}).  
Furthermore, because HTML is a single semi-structured standard used across publishers, we could write relatively generic HTML-processing algorithms applicable to content published across journals.  
Our focus on using HTML limits us to relatively newer articles - typically those published after 1996 - because before this time most publications are only available as scanned PDF files.  
However, because the rate of publication across the field has grown exponentially, this HTML-available subset constitutes the majority of published neuroscience articles.

@@

When designing our processing algorithms, we parsed data strings from right to left: 
first searching for data entities contained within parentheses, then for entities contained to the right of the $\pm$ term, and finally the remaining term which we assumed to refer to the mean term.  
We found that occasionally data were reported as ``XX (LL - HH)'' -- where LL and HH indicate the lower and upper limits of a data range -- and accounted for these cases similarly.  
We used regular expressions to identify entities such as digits, decimal signs, parentheses, and $\pm$ signs.  
We then converted the individual data elements which were encoded as textual strings of digits to double precision decimal entities before storing these into our database.  
Our focus here was primarily on parsing the mean value from a data record (i.e., summarizing the properties of a number of recorded neurons), but also extracted and stored the error term and sample size where possible.  
Using these methods, we were able to extract 2176 electrophysiological values for 93 distinct neuron types within 279 articles.

@@

\subsection{Object models and relational database}
We stored extracted data and metadata using a relational database implemented in MySQL (\url{http://dev.mysql.com/doc/refman/5.6/en/}) built from a Python Django object model (\url{https://www.djangoproject.com/}).  
The object model contains classes for a number of fields, such as full article texts, electrophysiological properties, neuron types, synonyms, electrophysiological data values, and experimental metadata (Fig. S1).  
A useful feature of the relational nature of the database is that it enables linking between classes (e.g., linking between neuron types and electrophysiological properties reported by a single investigator across multiple articles). 
This linking feature facilitates efficient and arbitrary querying of data; 
for example, querying for known electrophysiological data on olfactory bulb mitral cells recorded \emph{in vitro} and published between the dates 2000 and 2004.  
For example, such a feature could be used to assess whether measurements of olfactory bulb mitral cells have changed as a function of time or are dependent upon whether the data are collected \emph{in vitro} or \emph{in vivo}.

@@

\begin{itemize}
\item \url{http://neuroelectro.org/api/1/n/} : Returns a list of all neurons with electrophysiological data indexed in NeuroElectro.
\item \url{http://neuroelectro.org/api/1/nedm/?nlex=sao830368389} : Returns a list of all indexed data on CA1 pyramidal cells (queried using the NeuroLex identifier for CA1 pyramidal cells, \emph{sao830368389}).  
\item \url{http://neuroelectro.org/api/1/nes/?e__name=Input%20Resistance&n__name=Cerebellum%20Purkinje%20Cell} : Returns a data record composed of the mean, standard deviation, and sample size n, summarizing input resistance measurements from cerebellar Purkinje cells based on all indexed articles in NeuroElectro database.  
Here the database query is performed using the textual strings for the electrophysiological and neuron type concepts.
\end{itemize}

@@

Given our laboratory's focus on olfactory circuits, we conducted a natural experiment to compare the efficacy of biophysical property extraction using these semi-automated methods versus traditional methods which do not make use of algorithmic text-mining as a pre-processing step.  
In a seven-hour curation session (evoking the classic American parable of John Henry versus the steam-powered hammer), a senior graduate student in our laboratory identified 91 electrophysiological measurements (focusing on resting membrane potential, input resistance, membrane time constant, spike amplitude, spike width, and spike threshold) from 35 articles for 7 olfactory bulb neuron types using only prior knowledge of which articles and investigators were likely to have reported such electrophysiological data.  
In a comparable seven-hour curation session using our semi-automated methods, a single curator (with similar expertise to the first curator) identified 551 electrophysiological measurements from 70 articles across 40 neuron types throughout the nervous system.  
Moreover, this comparison would likely tilt even more in favor of the semi-automated methods had the curators been less familiar with the primary literature.

\subsection{Scalability of current approach}
We note that multiple steps in our approach require manual intervention by an expert curator in order for electrophysiological measurements to be extracted with an acceptably low error rate.  
Namely, an expert curator needs to confirm which of the machine-identified candidate neuron types are recorded from in each article and where data from the normotypic or ``control'' states of these neurons are textually referenced within a data table.  
Moreover, given the current accuracy of the unsurpervised algorithmic assignment of electrophysiological concepts and experimental metadata (78\% and 50\%, respectively), these also need to be manually validated and corrected and normalized as required by an expert.  
Given the necessity of these manual steps, the scalability of our current approach is limited by our ability to manually curate this information or by our ability to improve the error rate of the automated methods.  
Despite this limitation, our current pipeline is still much faster than a purely manual one.  The methodology could be further improved by addressing falsely matching entities (such as EPSP amplitude in section \ref{sec:ephys_extraction}) by either auto-suggesting alternative matches returned by string-similarity algorithms or by simply adding these concepts to the electrophysiolgical ontology.  
Moreover, these improvements would facilitate formally computing the sensitivity and specificity of these entity recognition methods.

\subsection{Preliminary use of NeuroElectro in scientific work}
The NeuroElectro project is intended to facilitate scientific investigation by providing easy access to large quantities of data about neurons.  
Because the data is machine-readable, we have already begun to conduct several analyses that would not be possible without this resource.  
First, we have begun an investigation of the relationships between neurons as defined by the similarity of their electrophysiological properties.  
This information can be used to make predictions about as yet unmeasured properties.  
Second, we have begun to explore the relationship between patterns of gene expression (using both the Allen Brain Atlas \citep{lein_genome-wide_2007} and single cell qPCR approaches) and electrophysiological properties of neurons.  
Third, we have begun automated testing of quantitative neuron models in concert with SciUnit \citep{omar_sciunit_2014}, under the reasonable assumption that these models should be constrained by the available experimental data.  
These projects are described in manuscripts currently in preparation.

\subsection{Extensions and improvements to the current semi-automated algorithms}
Currently, neuron type identification is a critical bottleneck in our approach.  
One potential improvement would be to replace the nonspecific bag-of-words approach we are currently using in favor of a bioNLP classifier-based approach \citep{mccallum_mallet:_2002}.  
Specifically, we propose adapting the named entity recognition methodology used by the WhiteText project for tagging brain regions mentioned in literature \citep{french_automated_2009,french_using_2012} and first identifying spans of text likely to pertain to a neuron type before mapping these textual spans to a individual neuron type within the neuron ontology.

@@

Given the relatively low accuracy of the automated approach to identifying neuron types, there may be several avenues through which this process can be improved. 
For example, we note the automated approach was particularly ineffective when the neuron type investigated within an article was not already described in NeuroLex or when the neuron had an insufficient list of synonyms associated with it.  
The current implementation of NeuroElectro also does not consider common neuron type acronyms (e.g., that olfactory bulb mitral cells are commonly referred to as ``MCs'').  
Adding acronym and abbreviation identification to future iterations will thus likely improve the automated approach \citep{okazaki_building_2006,french_using_2012}.  
Moreover, our current implementation of the bag-of-words algorithm would likely be enhanced via minor improvements, such as only identifying neurons using the text of the abstract or results and discarding text from the introduction or discussion. 
As neuron identification forms the major bottleneck in the scalability of NeuroElectro due to the requirement for manual curation, we plan to address this bottleneck in future revisions.

@@

Future developments in machine extraction of data from the scientific literature will be of great benefit.  
These should include better semantic understanding of context, ranging from relatively unambiguous notations such as units, to syntax-parsing of free-form prose that relates objects of study to their reported properties.  
Much progress has been made by computer scientists in some of these areas, and more future engagement with their research should enable vastly more data to be extracted from the literature.

We believe that, if successful, the use of NeuroElectro will influence the practices of scientists writing papers and reporting results.  
Specifically, we recommend the usage of common standards and definitions for basic physiological measurements \citep{toledo-rodriguez_correlation_2004} and neuron types \citep{ascoli_petilla_2008,larson_neurolex.org:_2013}.  
Moreover, we advocate that, where possible, scientists report more basic physiological data overall and report such data using machine-parsable data tables.  
These recommendations could be made informally by journals (in particular, requested by reviewers during manuscript review) as well as by funding agencies.
This change would make it easier for scientists to find and make use of data collected by others.  
Such a culture shift has the potential to make science function more effectively and efficiently to facilitate discovery.

@@

\bibliographystyle{template/frontiersinSCNS&ENG} % for Science and Engineering articles
\bibliography{references}
\
\section*{Supplemental Data}  
\begin{figure*}[htbp]
\centering
\includegraphics[scale = .38, angle =90]{figures/django_models_v2.jpg}

Fig. S1: Illustration of NeuroElectro relational database schema
%\label{db_schema}
\end{figure*}
%SDB: anyway you can increase the font on at least some of these pathways? Or split into multiple figures showing subsections of the relations? As is, it is very difficult to look at this and take anything away from it.
